---
title: "Clinical Validation Study Example in the NIH *All of Us* Research Program Database"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Clinical Validation Study Example in the NIH *All of Us* Research Program Database}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warnings = FALSE, 
                      message = FALSE, 
                      comment = "#>")
```

**Author**: Patrick Wu  
**Date**: 2021-04-05   

**Note:** In this vignette, individual-level data are used for demonstration purposes only, and were not derived from real-world patients.  

## Set up packages
Install required packages
```r
install.packages("janitor")
devtools::install_github("pwatrick/DrugRepurposingToolKit")
```
Import packages
```{r}
suppressPackageStartupMessages({
  library(glue); 
  library(lubridate); 
  library(tidyverse);
  library(vroom);
  library(broom);
  library(DrugRepurposingToolKit)
}) 
set.seed(1)
```

## Download data
Define variables for functions
```r
drug_concept_id <- "1332418"
biomarker_concept_id <- "3004249"
indication_drug_concept_ids <-
        glue("({drug_concept_id},19020068,1309068,1345141,904542,19026595,
             40226742,1398937,1387219,932745,1335471,19011933,1334456,1318137,
             19011879,1351557,19062195,1308842,1338005,1400959,1319751,1366279,
             915829,1309799,19132932,948787,1335301,991382,1337068,902427,1319880,
             19026333,1346823,1312742,19049024,1346686,19054768,1317967,1373225,
             1367500,1354118,1326012,1331235,1361711,1366237,970250,1353776,1376289,
             1341927,19026800,1344992,1342439,1316354,19026714,1327978,19055337,1308216,
             974166,19126939,1322081,19132964,19063575,19012121,1350489,978555,19011908,
             904639,950370,987406,1319232,1395058,1317640,1363749,1347384,741530,1342001,
             1316262,1318853,1328165,1321636,19081667,1353766,942350,1313200,1363053,1305447,
             992590,956874,19020994,1310756,907013,19103548,1307046,19011340,1314002,1401581,
             1326378,1307863,1336926,1319998,19026551,1340128,19125119,1335539,905273,1332418,
             1345858,1327256,19026847,1373928,19125719,1341238,19055034,1386957,1362225,1314577,
             19011878)")
```
Download covariates data
```r
## Create SQL query
covariates_query <-
  DrugRepurposingToolKit::extract_clinical_data(drug_concept_id, 
                                                biomarker_concept_id, 
                                                table_name = "hypertension_amlodipine_drugs", 
                                                table_type = "covariates", 
                                                indication_drug_concept_ids)
## Execute query
hypertension_amlodipine_covariates <- 
  bigrquery::bq_table_download(
    bigrquery::bq_dataset_query(Sys.getenv("WORKSPACE_CDR"),
                                covariates_query,
                                billing=Sys.getenv("GOOGLE_PROJECT")))

## Save `hypertension_amlodipine_covariates` to CSV file
my_bucket <- Sys.getenv('WORKSPACE_BUCKET')
covariates_file <- "hypertension_amlodipine_covariates.csv"
write_csv(hypertension_amlodipine_covariates, covariates_file)
system(paste0("gsutil cp ./", covariates_file, " ", my_bucket, "/data/"), intern=T)
system(paste0("gsutil ls ", my_bucket, glue("/data/hypertension_amlodipine_covariates.csv")), intern=T)
```
Example `hypertension_amlodipine_covariates` table
```{r echo=FALSE}
covariates_table <- dplyr::tribble(
  ~person_id, ~dob, ~gender_concept_id, ~race_concept_id, ~drug_concept_id, ~first_drug_exposure, ~last_drug_exposure, 
  1, '1957-04-18', 8532, 38003598, 1332418, '2010-03-02', '2010-12-07', 
  2, '1985-05-01', 8507, 8527, 1332418, '2016-09-30	', '2016-11-10', 
  3, '1950-12-25', 8507, 8527, 1332418, '2011-01-24', '2014-08-01', 
)
DT::datatable(covariates_table, options = list(scrollX = TRUE))
```

Download drugs data
```r
## Create SQL query
drugs_query <-
  DrugRepurposingToolKit::extract_clinical_data(drug_concept_id, 
                                                biomarker_concept_id, 
                                                table_name = "hypertension_amlodipine_drugs", 
                                                table_type = "drugs", 
                                                indication_drug_concept_ids)
## Execute query
hypertension_amlodipine_drugs <- 
  bigrquery::bq_table_download(
    bigrquery::bq_dataset_query(Sys.getenv("WORKSPACE_CDR"),
                                drugs_query,
                                billing=Sys.getenv("GOOGLE_PROJECT")))

## Save `hypertension_amlodipine_drugs` to CSV file
my_bucket <- Sys.getenv('WORKSPACE_BUCKET')
drugs_file <- "hypertension_amlodipine_drugs.csv"
write_csv(hypertension_amlodipine_drugs, drugs_file)
system(paste0("gsutil cp ./", drugs_file, " ", my_bucket, "/data/"), intern=T)
system(paste0("gsutil ls ", my_bucket, glue("/data/hypertension_amlodipine_drugs.csv")), intern=T)
```

Example `hypertension_amlodipine_drugs` table
```{r echo=FALSE}
drugs_table <- dplyr::tribble(
  ~person_id, ~start_date, ~end_date, ~drug_exposure_start_date, ~drug_concept_id, 
  1, '2009-03-02', '2011-03-02', '2010-03-02', 1332421, 
  2, '2015-09-30', '2017-09-30', '2016-11-09', 19103475,  
  3, '2010-01-24', '2012-01-24', '2011-01-24', 957138, 
)
DT::datatable(drugs_table, options = list(scrollX = TRUE))
```

Download biomarkers data
```r
## Create SQL query
biomarkers_query <-
  DrugRepurposingToolKit::extract_clinical_data(drug_concept_id, 
                                                biomarker_concept_id, 
                                                table_name = "hypertension_amlodipine_drugs", 
                                                table_type = "biomarkers", 
                                                indication_drug_concept_ids)
## Execute query
hypertension_amlodipine_biomarkers <- 
  bigrquery::bq_table_download(
    bigrquery::bq_dataset_query(Sys.getenv("WORKSPACE_CDR"),
                                biomarkers_query,
                                billing=Sys.getenv("GOOGLE_PROJECT")))

## Save `hypertension_amlodipine_covariates` to CSV file
my_bucket <- Sys.getenv('WORKSPACE_BUCKET')
biomarkers_file <- "hypertension_amlodipine_biomarkers.csv"
write_csv(hypertension_amlodipine_biomarkers, biomarkers_file)
system(paste0("gsutil cp ./", biomarkers_file, " ", my_bucket, "/data/"), intern=T)
system(paste0("gsutil ls ", my_bucket, glue("/data/hypertension_amlodipine_biomarkers.csv")), intern=T)
```

Example `hypertension_amlodipine_biomarkers` table
```{r echo=FALSE}
biomarkers_table <- dplyr::tribble(
  ~person_id, ~start_date, ~measurement_date, ~end_date, ~measurement_concept_id, ~value_as_number,
  1, '2009-03-02', '2010-03-02', '2011-03-02', 3004249, 145, 
  2, '2015-09-30', '2017-06-07', '2017-09-30', 3004249, 153, 
  3, '2010-01-24', '2011-01-24', '2012-01-24', 3004249, 135, 
)
DT::datatable(biomarkers_table, options = list(scrollX = TRUE))
```

## Process data
Process data to create table to perform paired two-tailed t-test
```r
#Import data from saved CSV files
r_c <- vroom::vroom("hypertension_amlodipine_covariates.csv",
                    .name = janitor::make_clean_names, 
                    col_types = cols())
r_d <- vroom::vroom("hypertension_amlodipine_drugs.csv",
                    .name = janitor::make_clean_names, 
                    col_types = cols())
r_b <- vroom::vroom("hypertension_amlodipine_biomarkers.csv",
                    .name = janitor::make_clean_names, 
                    col_types = cols())


#Prepare data for paired two-tailed t-test
drug <- "amlodipine"
phenotype <- "Hypertension"
biomarker <- "Systolic Blood Pressure"
indication_drugs <- DrugRepurposingToolKit::drugsHypertension
concept_id_exclusions <- c(2212451)

htn_amlodipine_processed_data <- 
  DrugRepurposingToolKit::clean_process_clinical_data(
    drug, phenotype, biomarker, indication_drugs, concept_id_exclusions, 
    r_c, r_d, r_b) %>% 
  mutate(drug = "amlodipine")
```

Example `htn_amlodipine_processed_data` table
```{r echo = FALSE}
procssed_data_table <- dplyr::tribble(
  ~person_id, ~gender_concept_id, ~race_concept_id, ~start_date, ~first_drug_exposure, ~final_end_date, ~age, ~biomarker_baseline_value, ~biomarker_treatment_value, ~biomarker_change, ~biomarker_name, ~drug, 
  1, 8532, 38003598, '2009-03-02', '2010-03-02', '2010-12-07', 51.87, 143, 138, -5, "Systolic Blood Pressure", "amlodipine",
  2, 8507, 8527, '2015-09-30', '2016-09-30	', '2016-11-10', 30.41, 140, 140, 0, "Systolic Blood Presure", "amlodipine",  
)
DT::datatable(procssed_data_table, options = list(scrollX = TRUE))
```

## Run hypothesis test
```r
#Run paired two-tailed t-test
htest <- t.test(htn_amlodipine_processed_data$biomarker_treatment_value, 
                htn_amlodipine_processed_data$biomarker_baseline_value, 
                conf.level = 0.95, 
                paired = TRUE)
tidy_htest <- tidy(htest) %>% 
    mutate(drug = drug, 
           se = htest$stderr,
           biomarker = "Systolic Blood Pressure", 
           phenotype = "Hypertension", 
           source = "All of Us") %>% 
    select(source, drug, phenotype, biomarker, estimate, se, statistic, 
           p.value, parameter, conf.low, conf.high, method, alternative)

tidy_htest
```

Example `tidy_htest` table
```{r echo = FALSE}
example_htest <- dplyr::tribble(
  ~source, ~drug, ~phenotype, ~biomarker, ~estimate, ~se, ~statistic, ~p.value, ~parameter, ~conf.low, ~conf.high, ~method, ~alternative, 
  "All of Us", "amlodipine", "Hypertension", "Systolic Blood Pressure", -5.24, 0.52, -9.99, 3.92e-22, 733, -6.27, -4.21, "Paired t-test", "two.sided", 
)
DT::datatable(example_htest, options = list(scrollX = TRUE))
```

## Metadata

Workspace:  

  - Name: Drug Repurposing Validation Study  
  - Owner: patrickwu@researchallofus.org  
  
Dataset:  

  - All of Us Dataset v4  
  - Data access level: Registered  

Cloud compute profile:  

  - CPUs: 4  
  - RAM (GB): 15  
  - Disk (GB): 100  
  - Compute type: Standard VM  
  
Time and cost to run notebook (without installing packages):  

  - Time: < 1 hour  
  - Cost: < $0.20 USD  
